Tree height and hydraulic traits shape growth responses across droughts in a temperate broadleaf forest

### 2020-07-16 manuscript/response_to_reviews.Rmd

Table 3: please, add also the number of trees, it is more informative than the number of cores to check sample size in your study.

**In this study, we have one core per tree, now specified in the methods. We have changed the table header to "n trees".**

Table 3: the authors could show some test to see if the traits presented are different among species.

**We now show differences in traits among species in Table S4.**

*need to add ANOVA, prsent reults in paper*

@@

I enjoyed the paper very much. It’s clearly and succinctly written, and brings together a range of complementary datasets to address some very topical questions in forest ecology. I could definitely see the paper being of interest to a broad readership. However, I do have some important questions and suggestions about the analysis which I think need more careful thought. Below I’ve listed my main concerns, following which there are some other comments that the authors may want to take into consideration in revising their paper.

Main comments

L139-140: makes sense to not core live trees in the plot (surprised you got permission to do this in 2011!). But doesn’t this introduce a bias in your analysis? Trees that died may have had very different responses to drought that those which survived. Do you get similar results if you analyse trees that lived and died separately?

**We agree that this could be a concern, but do not believe it is applicable here, for two reasons. First, this would be more likely to introduce a bias if the trees cored dead were killed by drought. We are aware of studies showing differences in climate sensitivity of trees killed by drought versus those that survived. However,these trees were not killed by drought; there was no significant drought in the years in which dead trees were cored or in the 3 previous years. Second, quoting our previous publication using these cores (Helcoski et al. 2019): " Chronologies of trees cored live and dead were pooled following analyses showing similar climate sensitivity at least up to 2009 (that is excluding 7–8 yr before death) for all four species with ≥ 10 cores in each category (LITU, QURU, QUVE, FRAM; comparison plots available at https://github.c om/SCBI-ForestGEO/climate_sensitivity_cores/tree/master/ results/live_vs_dead).".**

**To clarify that this shouldn't be an issue, we have added the following sentence (methods section): "We note that drought was probably not a cause of mortality for these trees, as monthly May-Aug $PDSI$ did not drop below -1.75 in these years or the three years prior (2013-2017), and that trees cored dead displayed similar climate sensitivity to trees cored live [@helcoski_growing_2019]."**

### 2020-07-16 manuscript/response_to_reviews.Rmd

**We have gone through the manuscript / tables and figures and updated where it said "dAICc" to be $\Delta$AICc. **

Following the two previous comments, I suggest to leave in Table 5 just the best model (rather that a multimodel inference as is presented now) to discuss more clearly the covariates included. And I suggest discussing only the model with all years together. This would be more robust and avoid ramble differences between years that mostly look spurious and with a difficult physiological explanation. This would help to ensure a better use of the Resilience indices.

**In the spirit of this comment (simplifying presentation), we have created a new figure (*Fig. 4*) visualizing the top model results. We moved Table 5 to the Appendix (now Table S6, plus a parallel Table S7 for $Rt_{ARIMA}$).**

*We do want to show the three separate drought periods, as we feel that this makes the analysis richer and helps to separate real drivers from spurious correlations. However, in order to make models more consistent across years and reduce spurious correlations, we now include the criteria that for a trait to be included as a candidate variable in the full model, the sign of coefficients must be consistent across all droughts (in addition to meeting the AIC criteria, as before). This resulted in only two candidate variables for the main model ($PLA_{dry}$ and $\pi_{tlp}$). (Removed was ring porosity, which had opposite effects in different droughts.) *

### 2020-07-16 manuscript/climate_sensitivity_draft.Rmd
worked on reviewer responses
removed CP from full model, focus on just main results
issue #111 , #112

- Individual-level drought resistance decreased with tree height, which was overall a stronger predictor than crown exposure. The potentially greater rooting volume of larger trees did not confer an advantage in sites with low topographic wetness index. Resistance was greater among species whose leaves experienced less shrinkage upon desiccation and lost turgor (wilted) at more negative water potentials.

@@

Here, we examine how tree height, microenvironment characteristics, and species' traits collectively shape drought responses. 
We test a series of hypotheses and associated specific predictions (Table 1) based on the combination of tree-ring records from three droughts (1966, 1977, 1999), species functional and hydraulic trait measurements, and census data from a large forest dynamics plot in Virginia, USA (Table 2). 
First, we focus on the role of tree height and its interaction with microenvironment. 
We test hypotheses designed to disentangle the relative importance of tree height; crown exposure; and soil water availability, which should be greater for larger trees in dry but not in perpetually wet microsites. 
Second, we focus on the role of species' functional and hydraulic traits, testing the hypothesis that species’ traits-–particularly leaf hydraulic traits-–predict $Rt$. 
We test predictions that drought resistance is higher in ring-porous than semi-ring and diffuse-porous species, that it is correlated with wood density--either postively [@greenwood_tree_2017] or negatively [@hoffmann_hydraulic_2011]-- and positively correlated with $LMA$, and that hydraulic leaf traits including $PLA_{dry}$ and $\pi_{tlp}$ are better predictors.

### Materials and Methods
*Study site*

@@

@@

We analyzed tree-ring data (xylem growth increment) from 571 trees representing the twelve species with the greatest contributions to woody aboveground net primary productivity ($ANPP_{stem}$), which together comprised 97% of study plot $ANPP_{stem}$ between 2008 and 2013 [@helcoski_growing_2019] (Table 3; Fig. **S1**). Cores (one per tree) were collected within the ForestGEO plot at breast height (1.3m) in 2010-2011 or 2016-2017. In 2010-2011, cores were collected from randomly selected live trees of each species that had at least 30 individuals $\ge$ 10 cm DBH [@bourg_initial_2013]. In 2016-2017, cores were collected from all trees found dead during annual mortality censuses [@gonzalezakre_patterns_2016]. Cores were sanded, measured, and crossdated using standard procedures, as detailed in [@helcoski_growing_2019]. 
The resulting chronologies (Fig. **1a**) were published in Zenodo (DOI: 10.5281/zenodo.2649302) in association with @helcoski_growing_2019.

@@

To characterize how environmental conditions vary with height, data were obtained from the NEON tower located <1km from the study area via the neonUtilities package [@R-neonUtilities]. We used wind speed, relative humidity, and air temperature data, all measured over a vertical profile spanning heights from 7.2 m to above the top of the tree canopy (31.0 or 51.8m, depending on censor), for the years 2016-2018 [@neon_national_2018]. After filtering for missing and outlier values, we determined the daily minima and maxima, which we then aggregated at the monthly scale.

@@

@@

*Tree size, microenvironment, and drought resistance*

@@

Crown position varied as expected with height (dominant > co-dominant > intermediate > suppressed), but with substantial variation (Fig. **3d**). 
Crown position was not included in any of the top models for all droughts or for 1966 individually (Table 1). 
It did, however, come out in the best models for 1977 and 1999 (Fig. **4**; Table S6).
When $Rt_{ARIMA}$ was used as the response variable, crown position was never in a best model, and was included among the top models only for 1977 (Table S7).
Crown position did not have a consistent influence: dominant trees had the lowest $Rt$ (or $Rt_{ARIMA}$) in 1977, but in 1999 intermediate and suppressed trees had lower $Rt$ (Fig. 4; Tables S6-S7).

>>>>

- Individual-level drought resistance decreased with tree height, which was strongly correlated with crown exposure. The potentially greater rooting volume of larger trees did not confer an advantage in sites with low topographic wetness index. Resistance was greater among species whose leaves experienced less shrinkage upon desiccation and lost turgor (wilted) at more negative water potentials.

@@

Here, we examine how tree height, microenvironment characteristics, and species' traits collectively shape drought responses. 
We test a series of hypotheses and associated specific predictions (Table 1) based on the combination of tree-ring records from three droughts (1966, 1977, 1999), species functional and hydraulic trait measurements, and census data from a large forest dynamics plot in Virginia, USA (Table 2). 
First, we focus on the role of tree size and its interaction with microenvironment. 
We examine the contemporary relationship between tree height and microenvironment, including growing season meteorological conditions and crown exposure. 
We then test whether, consistent with most forests globally, larger-diameter, taller trees tend to have lower drought resistance ($Rt$) in this forest, which is in a region (eastern North America) represented by only two studies in the global review of Bennett et al. (2015).
We also test for an influence of potential access to available soil water, which should be greater for larger trees in dry but not in perpetually wet microsites. 
Finally, we focus on the role of species' functional and hydraulic traits, testing the hypothesis that species’ traits-–particularly leaf hydraulic traits-–predict $Rt$. 
We test predictions that drought resistance is higher in ring-porous than semi-ring and diffuse-porous species, that it is correlated with wood density--either postively [@greenwood_tree_2017] or negatively [@hoffmann_hydraulic_2011]-- and positively correlated with $LMA$, and that hydraulic leaf traits including $PLA_{dry}$ and $\pi_{tlp}$ are better predictors.

### Materials and Methods
*Study site and microclimate*

@@

*To characterize how environmental conditions vary with height, data were obtained from the NEON tower located <1km from the study area via the neonUtilities package [@R-neonUtilities]. We used wind speed, relative humidity, and air temperature data, all measured over a vertical profile spanning heights from 7.2 m to above the top of the tree canopy (31.0 or 51.8m, depending on censor), for the years 2016-2018 [@neon_national_2018]. After filtering for missing and outlier values, we determined the daily minima and maxima, which we then aggregated at the monthly scale.* **(work on organization of this.)**

@@

We analyzed tree-ring data (xylem growth increment) from 571 trees representing the twelve species with the greatest contributions to woody aboveground net primary productivity ($ANPP_{stem}$), which together comprised 97% of study plot $ANPP_{stem}$ between 2008 and 2013 [@helcoski_growing_2019] (Table 3; Fig. **S1**). Cores (one per tree) were collected within the ForestGEO plot at breast height (1.3m) in 2010-2011 or 2016-2017. In 2010-2011, cores were collected from randomly selected live trees of each species that had at least 30 individuals $\ge$ 10 cm DBH [@bourg_initial_2013]. 
In 2016-2017, cores were collected from all trees found dead during annual mortality censuses [@gonzalezakre_patterns_2016]. 
We note that drought was probably not a cause of mortality for these trees, as monthly May-Aug $PDSI$ did not drop below -1.75 in these years or the three years prior (2013-2017), and that trees cored dead displayed similar climate sensitivity to trees cored live [@helcoski_growing_2019]. 
Cores were sanded, measured, and crossdated using standard procedures, as detailed in [@helcoski_growing_2019]. 
The resulting chronologies (Fig. **1a**) were published in Zenodo (DOI: 10.5281/zenodo.2649302) in association with @helcoski_growing_2019.

@@

@@

*Tree height and microenvironment*

Crown position varied as expected with height (dominant > co-dominant > intermediate > suppressed), but with substantial variation (Fig. **3d**). 
Correlation between height and canopy position was 0.73.

@@

*Tree height, microenvironment, and drought resistance*

@@

### 2020-07-16 manuscript/tables&figures.Rmd
removed CP from full model, focus on just main results
issue #111 , #112

\blandscape
```{r Table 1, eval = TRUE, echo=FALSE, warning=FALSE}
library(knitr)
library(kableExtra)
hypothesis_table <- read.csv("tables_figures/publication/table1_hypothesis_table.csv", stringsAsFactors = FALSE, check.names = FALSE)
kable(hypothesis_table, format = "latex", booktabs = TRUE, caption = "Table 1. Summary of hypotheses, corresponding specific predictions, and results.") %>%
  kable_styling(latex_options = c("scale_down", "hold_position"), font_size = 8.5, protect_latex = TRUE) %>%
  pack_rows("Tree size and microenvironment", 1, 4, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
  pack_rows("Taller height makes trees less drought resistant.", 1, 1, latex_gap_space = "1em", colnum = 1, hline_before = FALSE, bold = FALSE, italic = TRUE) %>%
  pack_rows("Trees with more exposed crowns have lower drought resistance.", 2, 3, latex_gap_space = "1em", colnum = 1, hline_before = FALSE, bold = FALSE, italic = TRUE) %>%
  pack_rows("Small trees (lower root volume) in drier microhabitats have lower drought resistance.", 4, 4, latex_gap_space = "1em", colnum = 1, hline_before = FALSE, bold = FALSE, italic = TRUE) %>%
  pack_rows("Species traits", 5, 9, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
  pack_rows("Species' traits--particularly leaf hydraulic traits--predict drought reistance.", 5, 9, latex_gap_space = "1em", colnum = 1, hline_before = FALSE, bold = FALSE, italic = TRUE) %>%
  add_header_above(c(" ", "Prediction supported?" = 4, " ")) %>%
  column_spec(1, width = "12cm") %>%
  column_spec(2:5, width = "1.2cm") %>%
  column_spec(6, width = "2cm") 
 # kableExtra::landscape()
```

Parentheses indicate that the prediction was supported by one but not all of the top  models (Table S6). Dash symbols indicate that the response was not significant (Table S4)/ not represented in any of the top models (Table S6).
Asterisks indicate that the $Rt_{ARIMA}$ models (Table S5, S7) yielded a different result: 
*$Rt_{ARIMA}$ had a difference in significance/ inclusion in top  models.
**$Rt_{ARIMA}$ model had a significant trend in the opposite direction of the $Rt$ model (if significant) or the prediction (if no significant trend in $Rt$ model).

@@

**Figure 1. Climate and species-level growth responses over our study period, highlighting the three focal drougths (a) and community-wide responses** Time series plot (**a**) shows peak growing season (May-August) climate conditions and residual chronologies for each species (see Table 3 for codes). PET and PRE data were obtained from the Climatic Research Unit high-resolution gridded dataset (CRU TS v.4.01; Harris et al. 2014). Focal droughts are indicated by dashed lines, and shading indicates the pre-drought period used in calculations of the resistance metric. Figure modified from Helcoski *et al.* (2019). Density plots (**b**) show the distribution of resistance values for each drought.

**Figure 2. Drought resistance, $Rt$, across species for the three focal droughts.** Drought are color coded as in Fig. 1. Species codes are given in Table 3.

**Figure 3. Height profiles in growing season climatic conditions and tree heights by crown position** Shown are averages ($\pm$ SD) of daily maxima and minima of (**a**) wind speed, (**b**) relative humidity ($RH$), and (**c**) air temperature ($T_{air}$) averaged over each month of the peak growing season (May-August) from 2016-2018. In these plots, heights are slightly offset for visualization purposes. Also shown is (**d**) 2018 tree heights by crown position (see Table 2 for codes). In all plots, the dashed horizontal line indicates the 95th percentile of tree heigts in the ForestGEO plot.

@@

![**Figure 1. Climate and species-level growth responses over our study period, highlighting the three focal drougths (a) and community-wide responses** Time series plot (a) shows peak growing season (May-August) climate conditions and residual chronologies for each species (see Table 3 for codes). PET and PRE data were obtained from the Climatic Research Unit high-resolution gridded dataset (CRU TS v.4.01; Harris et al. 2014). Focal droughts are indicated by dashed lines, and shading indicates the pre-drought period used in calculations of the resistance metric. Figure modified from Helcoski *et al.* (2019). Density plots (b) show the distribution of resistance values for each drought.](tables_figures/publication/Figure1.png){width=500px}

![**Figure 2. Drought resistance, $Rt$, across species for the three focal droughts.** Drought are color coded as in Fig. 1. Species codes are given in Table 3.](tables_figures/publication/Figure2_rt_across_sp.png){width=500px}

![**Figure 3. Height profiles in growing season climatic conditions and tree heights by crown position** The top row shows averages ($\pm$ SD) of daily maxima and minima of (a) wind speed, (b) relative humidity ($RH$), and (c) air temperature ($T_{air}$) averaged over each month of the peak growing season (May-August) from 2016-2018. In these plots, heights are slightly offset for visualization purposes. Also shown is (d) 2018 tree heights by crown position (see Table 2 for codes). In all plots, the dashed horizontal line indicates the 95th percentile of tree heigts in the ForestGEO plot.](tables_figures/publication/Figure3_NEON_vars_height_profile.png){width=500px}

>>>>

\blandscape
```{r Table 1, eval = TRUE, echo=FALSE, warning=FALSE}
library(knitr)
library(kableExtra)
hypothesis_table <- read.csv("tables_figures/publication/table1_hypothesis_table.csv", stringsAsFactors = FALSE, check.names = FALSE)
kable(hypothesis_table, format = "latex", booktabs = TRUE, caption = "Table 1. Summary of hypotheses, corresponding specific predictions, and results.") %>%
  kable_styling(latex_options = c("scale_down", "hold_position"), font_size = 8.5, protect_latex = TRUE) %>%
  pack_rows("Tree size and microenvironment", 1, 4, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
 pack_rows("Trees with more exposed crowns have lower drought resistance.", 1, 2, latex_gap_space = "1em", colnum = 1, hline_before = FALSE, bold = FALSE, italic = TRUE) %>%
  pack_rows("Taller trees are less drought resistant.", 3, 3, latex_gap_space = "1em", colnum = 1, hline_before = FALSE, bold = FALSE, italic = TRUE) %>%
  pack_rows("Small trees (lower root volume) in drier microhabitats have lower drought resistance.", 4, 4, latex_gap_space = "1em", colnum = 1, hline_before = FALSE, bold = FALSE, italic = TRUE) %>%
  pack_rows("Species traits", 5, 9, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
  pack_rows("Species' traits--particularly leaf hydraulic traits--predict drought reistance.", 5, 9, latex_gap_space = "1em", colnum = 1, hline_before = FALSE, bold = FALSE, italic = TRUE) 
  #add_header_above(c(" ", "Prediction supported?" = 4, " ")) %>%
  #column_spec(1, width = "12cm") %>%
  #column_spec(2:5, width = "1.2cm") %>%
  #column_spec(6, width = "2cm") 
 # kableExtra::landscape()
```

Parentheses indicate that the prediction was supported by one but not all of the top  models (Table S6). Dash symbols indicate that the response was not significant (Table S4), or not represented in any of the top models (Table S6).

@@

**Figure 1. Height profiles in growing season climatic conditions and tree heights by crown position** Shown are averages ($\pm$ SD) of daily maxima and minima of (**a**) wind speed, (**b**) relative humidity ($RH$), and (**c**) air temperature ($T_{air}$) averaged over each month of the peak growing season (May-August) from 2016-2018. In these plots, heights are slightly offset for visualization purposes. Also shown is (**d**) 2018 tree heights by crown position (see Table 2 for codes). In all plots, the dashed horizontal line indicates the 95th percentile of tree heigts in the ForestGEO plot.

**Figure 2. Climate and species-level growth responses over our study period, highlighting the three focal drougths (a) and community-wide responses** Time series plot (**a**) shows peak growing season (May-August) climate conditions and residual chronologies for each species (see Table 3 for codes). PET and PRE data were obtained from the Climatic Research Unit high-resolution gridded dataset (CRU TS v.4.01; Harris et al. 2014). Focal droughts are indicated by dashed lines, and shading indicates the pre-drought period used in calculations of the resistance metric. Figure modified from Helcoski *et al.* (2019). Density plots (**b**) show the distribution of resistance values for each drought.

**Figure 3. Drought resistance, $Rt$, across species for the three focal droughts.** Drought are color coded as in Fig. 1. Species codes are given in Table 3.

@@

![**Figure 1. Height profiles in growing season climatic conditions and tree heights by crown position** The top row shows averages ($\pm$ SD) of daily maxima and minima of (a) wind speed, (b) relative humidity ($RH$), and (c) air temperature ($T_{air}$) averaged over each month of the peak growing season (May-August) from 2016-2018. In these plots, heights are slightly offset for visualization purposes. Also shown is (d) 2018 tree heights by crown position (see Table 2 for codes). In all plots, the dashed horizontal line indicates the 95th percentile of tree heigts in the ForestGEO plot.](tables_figures/publication/Figure3_NEON_vars_height_profile.png){width=500px}

![**Figure 2. Climate and species-level growth responses over our study period, highlighting the three focal drougths (a) and community-wide responses** Time series plot (a) shows peak growing season (May-August) climate conditions and residual chronologies for each species (see Table 3 for codes). PET and PRE data were obtained from the Climatic Research Unit high-resolution gridded dataset (CRU TS v.4.01; Harris et al. 2014). Focal droughts are indicated by dashed lines, and shading indicates the pre-drought period used in calculations of the resistance metric. Figure modified from Helcoski *et al.* (2019). Density plots (b) show the distribution of resistance values for each drought.](tables_figures/publication/Figure1.png){width=500px}

![**Figure 3. Drought resistance, $Rt$, across species for the three focal droughts.** Drought are color coded as in Fig. 1. Species codes are given in Table 3.](tables_figures/publication/Figure2_rt_across_sp.png){width=500px}
